### 2023-06-08 paper/localpsf.tex
dug up reference for O(k^2 N log N) operation count for H-matrix construction

We form $\mathbf{\Aker}_H$ using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package~\cite{Kriemann08}. For details about the algorithms used for geometrical clustering, H-matrix construction, and H-matrix operations in HLIBpro, we refer the reader to~\cite{BormGrasedyckHackbusch03,GrasedyckKriemannLeBorne08,Kriemann13}. Although $\mathbf{\Aker}$ is a dense $\fedim \times \fedim$ matrix, constructing $\mathbf{\Aker}_H$ only requires evaluation of $O(\hrank \fedim \log \fedim)$ kernel entries $\mathbf{\Aker}_{ij} = \widetilde{\Aker}(\fepoint_i, \fepoint_j)$, and these entries are computed via the radial basis function interpolation method described in Section~\ref{sec:approximate_kernel_entries}. Here $\hrank$ is the rank of the highest rank block in the H-matrix. 
We emphasize that the dense matrix $\mathbf{\Aker}$ is never formed. 
%We describe this H-matrix construction process in detail in Appendix~\ref{app:h_matrix}. 
%We form $\massmatrix_H$ using standard H-matrix methods for sparse matrices implemented within HLIBpro, using the same geometrical clustering block cluster tree as was used for $\mathbf{\Aker}$.

@@

\paragraph{(2) Building the H-matrix} Classical H-matrix construction techniques
%, which is described in Appendix~\ref{app:h_matrix}, 
require evaluating $O(\hrank \fedim \log \fedim)$ matrix entries of the approximation, where $\hrank$ is the H-matrix rank, i.e,  the maximum rank among the blocks of the H-matrix. To evaluate one matrix entry, first one must find the $\numnbr$ nearest sample points to a given point, where $\numnbr$ is the number of impulse responses used in the RBF interpolation. This is done using a precomputed kd-tree of sample points, and requires $O(\numnbr \log \nsamplepts)$ floating point and logical elementary operations. Second, one must find the mesh cells that the points $\{z_i\}_{i=1}^{\numnbr}$ reside in. This is done using an AABB tree of mesh cells, and requires $O(\numnbr \log \fedim)$ elementary operations. Third, one must evaluate finite element basis functions on each of those cells, which requires $O(\numnbr)$ elementary operations. Finally, one must solve a $\numnbr \times \numnbr$ linear system to perform the radial basis function interpolation, which requires $O(\numnbr^3)$ elementary operations. Therefore, building the H-matrix requires
%
%
\begin{equation*}
	O\left(\left(\hrank \fedim \log \fedim\right)\left(\numnbr \log \fedim + \numnbr^3\right)\right) \qquad \text{elementary operations}.
\end{equation*}
%
%
%This follows from the above discussion, and the fact that $\fedim \ge \nsamplepts$.

\paragraph{(3) Performing linear algebra operations with the H-matrix} It is well known that H-matrix methods for matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, matrix factorization, matrix inversion, and low rank updates require performing $O\left(\hrank^a \fedim \log(\fedim)^b\right)$
%\begin{equation}
%\label{eq:H_matrix_cost_operations}
%ASDASDAS	O(\hrank^2 \fedim \log(\fedim)^a) \qquad \text{elementary operations},
%\end{equation}
elementary operations, where $a,b \in \{0,1,2,3\}$ are constants which depend on the type of H-matrix used and the operation being performed~\cite{GrasedyckHackbusch03}\cite[Section 2.1]{Kriemann13}. We advise the reader to be aware that asymptotic complexity estimates for H-matrix operations can be a poor indicator of practical performance because of large prefactor constants, the complicated way in which recursive H-matrix algorithms map onto hardware, and difficulties taking advantage of parallelism. 
%In our experience with several software packages for H1 matrices, we have typically observed that matrix-vector products are extremely fast, matrix-matrix addition is fast, matrix LU factorization requires a comparatively moderate amount of time, factorized solves are extremely fast, and matrix-matrix multiplication and matrix inversion are comparatively slow. 
In the numerical results (Section~\ref{sec:numerical_results}), we use one matrix-matrix addition to add the H-matrix approximation of the data misfit term in the Hessian to the regularization term in the Hessian.
%Constructing an H-matrix representation of the regularization term in the Hessian requires one matrix inversion and two matrix-matrix multiplications. 
%Then, we use one matrix factorization to form a preconditioner for the Newton linear system. 
%We perform low-rank updates to the H-matrix as the Newton iterations proceed, as per the discussion in Appendix~\ref{sec:recycle_krylov}.
Symmetrizing $\mathbf{A}_H$ requires one matrix-matrix addition. Flipping negative eigenvalues to be positive requires a handful (typically around 5) matrix-matrix additions and matrix factorizations to factor the required shifted linear systems, and a number of factorized solves with these shifted linear systems that is proportional to the number of erroneous negative eigenvalues.
%semi-definite, described in Appendix~\ref{sec:make_spd}, is performed using two matrix-matrix additions, $\ratord+1$ matrix-matrix multiplications, and one matrix inversion, where $2^\ratord$ is the rational function order.

>>>>

We form $\mathbf{\Aker}_H$ using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package~\cite{Kriemann08}. For details about the algorithms used for geometrical clustering, H-matrix construction, and H-matrix operations in HLIBpro, we refer the reader to~\cite{BormGrasedyckHackbusch03,GrasedyckKriemannLeBorne08,Kriemann13}. Although $\mathbf{\Aker}$ is a dense $\fedim \times \fedim$ matrix, constructing $\mathbf{\Aker}_H$ only requires evaluation of $O(\hrank \fedim \log \fedim)$ kernel entries $\mathbf{\Aker}_{ij} = \widetilde{\Aker}(\fepoint_i, \fepoint_j)$ (see \cite{BebendorfRjasanow03}), and these entries are computed via the radial basis function interpolation method described in Section~\ref{sec:approximate_kernel_entries}. Here $\hrank$ is the rank of the highest rank block in the H-matrix. 
We emphasize that the dense matrix $\mathbf{\Aker}$ is never formed. 
%We describe this H-matrix construction process in detail in Appendix~\ref{app:h_matrix}. 
%We form $\massmatrix_H$ using standard H-matrix methods for sparse matrices implemented within HLIBpro, using the same geometrical clustering block cluster tree as was used for $\mathbf{\Aker}$.

@@

\paragraph{(2) Building the H-matrix} 
Classical H-matrix construction techniques
%, which is described in Appendix~\ref{app:h_matrix}, 
require evaluating $O(\hrank \fedim \log \fedim)$ matrix entries of the approximation, where $\hrank$ is the H-matrix rank, i.e,  the maximum rank among the blocks of the H-matrix \cite{BebendorfRjasanow03}. To evaluate one matrix entry, first one must find the $\numnbr$ nearest sample points to a given point, where $\numnbr$ is the number of impulse responses used in the RBF interpolation. This is done using a precomputed kd-tree of sample points, and requires $O(\numnbr \log \nsamplepts)$ floating point and logical elementary operations. Second, one must find the mesh cells that the points $\{z_i\}_{i=1}^{\numnbr}$ reside in. This is done using an AABB tree of mesh cells, and requires $O(\numnbr \log \fedim)$ elementary operations. Third, one must evaluate finite element basis functions on each of those cells, which requires $O(\numnbr)$ elementary operations. Finally, one must solve a $\numnbr \times \numnbr$ linear system to perform the radial basis function interpolation, which requires $O(\numnbr^3)$ elementary operations. Therefore, building the H-matrix requires
%
%
\begin{equation*}
	O\left(\left(\hrank \fedim \log \fedim\right)\left(\numnbr \log \fedim + \numnbr^3\right)\right) \qquad \text{elementary operations}.
\end{equation*}
%
%
%This follows from the above discussion, and the fact that $\fedim \ge \nsamplepts$.

\paragraph{(3) Performing linear algebra operations with the H-matrix} It is well known that H-matrix methods for matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, matrix factorization, matrix inversion, and low rank updates require performing $O\left(\hrank^a \fedim \log(\fedim)^b\right)$
%\begin{equation}
%\label{eq:H_matrix_cost_operations}
%ASDASDAS	O(\hrank^2 \fedim \log(\fedim)^a) \qquad \text{elementary operations},
%\end{equation}
elementary operations\footnote{We advise the reader to be aware that asymptotic complexity estimates for H-matrix operations can be a poor indicator of practical performance because of large prefactor constants, the complicated way in which recursive H-matrix algorithms map onto hardware, and difficulties taking advantage of parallelism.}, where $a,b \in \{0,1,2,3\}$ are constants which depend on the type of H-matrix used and the operation being performed~\cite{GrasedyckHackbusch03}\cite[Section 2.1]{Kriemann13}.
%In our experience with several software packages for H1 matrices, we have typically observed that matrix-vector products are extremely fast, matrix-matrix addition is fast, matrix LU factorization requires a comparatively moderate amount of time, factorized solves are extremely fast, and matrix-matrix multiplication and matrix inversion are comparatively slow. 
In the numerical results (Section~\ref{sec:numerical_results}), we use one matrix-matrix addition to add the H-matrix approximation of the data misfit term in the Hessian to the regularization term in the Hessian.
%Constructing an H-matrix representation of the regularization term in the Hessian requires one matrix inversion and two matrix-matrix multiplications. 
%Then, we use one matrix factorization to form a preconditioner for the Newton linear system. 
%We perform low-rank updates to the H-matrix as the Newton iterations proceed, as per the discussion in Appendix~\ref{sec:recycle_krylov}.
Symmetrizing $\mathbf{A}_H$ requires one matrix-matrix addition. Flipping negative eigenvalues to be positive requires a handful (typically around 5) matrix-matrix additions and matrix factorizations to factor the required shifted linear systems, and a number of factorized solves with these shifted linear systems that is proportional to the number of erroneous negative eigenvalues.
%semi-definite, described in Appendix~\ref{sec:make_spd}, is performed using two matrix-matrix additions, $\ratord+1$ matrix-matrix multiplications, and one matrix inversion, where $2^\ratord$ is the rational function order.
